prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-19 inst/MyArticle/paper.Rmd
incorporating suggestions from PN and CP.

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. However, identifying and quantifying intact proteins in complex samples containing hundreds of proteins is not possible using existing methods at the moment. Hence, the bottom-up mass spectrometric method is employed [@Bubis2017, @da2020philosopher]. Proteins that are extracted from biological samples are specifically and reproducible digested into protein fragments - peptides. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences[@eng2015deeper, @yu2016pipi, @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast, @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath, @demichev2020dia] are counted and aggregated to obtain precursor abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Protein quantification proteomics experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studied, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences of protein abundance between groups are significant. For this application it is important that the model provides an unbiased estimate of the false discovery rate (FDR) so that protein lists can be selected for follow-up experiments.   
Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins.

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analisis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

When analyzing parallel-group designs using a single explanatory variable, and contrasting the groups, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as $limma$ and $proDA$ allow us to fit a comprehensive variety of models and test various hypotheses; however, in-depth knowledge of design matrix specification is required [@law2020guide].

Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending,@MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarily to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

```{r tableOverview}
xx <- list(
  triqler = c("pd" = "Y"),
  MSstats = c("pd" = "Y", "rm" = "Y", "mem" = "y"),
  ROPECA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y", eb = "Y"),
  limma  = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y"),
  MSqRob2_rlm = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "?"),
  proDA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "Y"),
  prolfqua = c("pd" = "Y", "rm" = "Y", fd = "Y", int = "Y", mem = "Y",eb = "Y", md = "Y"))

bb <- data.frame(dplyr::bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, * - repeated measurements are modeled using a fixed effects. ? - the hurdle model was published but unclear if available in the msqrob2 package.")
```

We were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables to call various machine learning (ML) methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, many of these packages fit linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented (OO) design, with R linear models as a core, where features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")` or variance shrinkage could be easily enabled (see Figure \@ref(fig:ContrastUML)).

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the difference was imputed in such a way are explicitly annotated in the result table. 
Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020]. 
We show using a benchmark dataset that this, although trivial approach, can outperform methods based on data imputation or dropout models.

@@

We also implemented functionality to streamline the benchmarking of the implemented methods. We focus on the comparison of the various modeling approaches. We use the Ionstar [@shen2018ionstar] dataset to benchmark the analysis pipeline and modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`.

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models which should make it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes encapsulate the statistical modelling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (ContrastInterface), and there are various implementations allowing to perform differential expression analysis given linear or mixed effect models (Contrast), variance shrinkage (ContrastModerated), or impute contrasts in case of observations missing for an entire group of samples (ContrastSimpleImpute) can also be used to encapsulate and integrate differential expression analysis results of external tools such as SAINTExpress or proDA.

@@

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain various effect sizes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata].

[TODO : Describe robust scaling]

### Estimating contrasts for linear models

To allow for the comparison of different treatments, given a linear model $y = \beta X$, a contrast is computed by a linear combination $l$ of linear model parameters $\beta_m$. The difference estimate $\beta_c$, is given by:

\begin{eqnarray} 
\hat{\beta_{c}} &=& \sum l\beta_{m} 
\end{eqnarray}

with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of weights, defining the contrast $c$. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

[TODO : Describe how we produce linear combinations from model specification]

If there is no observation in a group, we assume that the protein was unobserved because the protein abundance is below the limit of detection (LOD) of the MS instrument. We estimate the detection limit as follows: if a protein is quantified just in a single sample of a treatment group, then it is at the instrument's detection limit. We compute the mean of those abundances and use it as the group mean if a protein is absent in a treatment group.

When computing differences among two groups $a$ and $b$, we will use the group mean $\bar{a}$ estimated from the data or the detection limit $a_d$ if no observations are present in the group. If for instance there are no observations in group $b$, then, $\Delta_{a,b} = \bar{a} - b_{dl}$ if $\bar{a} > b_{dl}$ or $\Delta_{a,b} = 0$ if $\bar{a} < b_{dl}$.

Furthermore, we assume that the variance of the protein is constant in all the conditions. To estimate the variance in the condition, we use the pooled variance based on all conditions:

Where, $n_g$ is the number of groups and $n$ is the number of observations.

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. [@Smyth2004linear] discuss how, using the empirical Bayes paradigm, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior $df$, and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:

\begin{equation} 
\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}
\end{equation}

and subsequently the $p$-values. We implemented this method in the class `ContrastModerated` (Figure \@ref(fig:ContrastUML)).

### Summarizing peptide level models

To obtain the median $p$-value of a protein, we first rescaled the peptide $p$-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

Afterwards, the median scaled $p$-value $\tilde{p}_s$ is determined and using the transformation below, transformed back onto the original scale.

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("lm", "linear model for peptides or proteins", "strategy_lm, Contrasts"),
    lmer = c("lmer" , "mixed effect models for peptides or proteins", "strategy_lmer, Contrasts"),
    gmi = c("group mean imputation", "impute missing group means", "ContrastsSimpleImpute"),
    ROPECA = c("ROPECA"," ", "ContrastsROPECA"),
    limma = c("limma" , "empirical Bayes variance shrinkage", "ContrastsModerated"),
    pic = c("SAINTexpress", "protein interaction scoring", "runSaint, ContrastsSaintExpress"),
    prodA = c("proDA", "dropout models", "strategy_proDA, Contrasts_proDa"), check.names =  FALSE)
xM <- data.frame(t(xM))

colnames(xM) <- c("id", "model", "prolfqua functions")
knitr::kable(tibble::as_tibble(xM[,1:3]), caption = "Models available in prolfqua.",
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down")

## Implementing an analysis workflow

The code snippets in this section demonstrate the basic usage of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. As input data we use subset of the Ionstar dataset containing $163$ proteins and $1258$ peptides. Peptide intensities are $\log_2$ transformed and robust z-score scaled but then rescaled to preserves the original range of the data using the method `robscale`. We show the distribution of the normalized intensities in Figure \@ref(fig:prepro) Panel A. Protein intensities are estimated using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. The boxplot (Panel D) shows the distribution of the protein intensity estimate in each condition. Finally, Panel C shows the distribution of the protein standard deviations for each condition.

(ref:scaling) Panel A - Peptide intensity distributions for 20 samples. For each sample a line with a different colour is shown. Panel B - Peptide intensities for protein _5NTD_ are shown using a line of different colour, and the protein intensity estimate is shown using a fat black line, Panel C - distribution of standard deviations of all proteins in each dilution group (a, b, c, d, e) and overall (all), Panel D - Distribution of protein intensities for protein _5NTD_.

The following code example illustrates how we compute contrast comparing the protein intensities in the condition. First, the differences and the linear model are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we examine the importance of the factors in the model using the ANOVA function, which facilitates model selection for models with one explanatory variable. Finally, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `addContrastResults` function merges both sets of contrast estimates, preferring the one obtained from the linear model. Figure \@ref(fig:exampleContrasts) Panel A shows the distribution of the $p$-values, while figure Panel B shows the volcano-plot for each comparison.

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among conditions
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()

## Benchmarking

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g., the false discovery proportion FDP. We use those plots to study how well linear, mixed effect models or $p$-value moderation models quantitative mass spectrometric high throughput experiments and compare them with results produced by `r BiocStyle::Biocpkg("MSstats")` [@MSstats2014] and `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020] (see also Figure \@ref(fig:FDRfdp))

The peptide intensities from the _MaxQuant_ _peptide.txt_ file were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e. proteins with a single peptide assignment. No further data filtering was used. We also did not use imputation. The protein level linear model was fitted to the protein intensities inferred from the peptide intensities using the Tukey's median polish. The mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model were fitted to the peptide intensities. The details of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` benchmark implementation are available online  [BenchmarkingIonstarData](https://wolski.github.io/prolfquabenchmark/articles/BenchmarkingIonstarData.html).

For benchmarking of `r BiocStyle::Biocpkg("MSstats")` we used the MSstats default parameters (for details see: [Benchmarking_MSstats](https://wolski.github.io/prolfquabenchmark/articles/))

@@

>>>>

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. Bottom-up mass spectrometric methods, where proteins are specifically and reproducible digested into protein fragments - peptides, are employed to identify and quantify proteins in complex samples containing hundreds to thousand of proteins, [@Bubis2017, @da2020philosopher]. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences[@eng2015deeper, @yu2016pipi, @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast, @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath, @demichev2020dia] are counted and aggregated to obtain precursor abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Proteomics quantification experiments enable monitoring relative abundances of thousands of proteins in biological samples. Most studies use parallel-group designs, where one or many treatment groups are compared to the control group [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with an increasing number of samples are studied, e.g., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether specific factors and factor interactions are significant, e.g., it can be tested if differences in protein abundance between groups are statistically significant.

An important aspect of mass spectrometric data are missing peptide and protein quantifications. [@rubin1976inference] classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the intensities of the observed features can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

When analyzing parallel-group designs using a single explanatory variable, and contrasting groups, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as $limma$ and $proDA$ allow us to fit a comprehensive variety of models and test various hypotheses; however, in-depth knowledge of design matrix specification using the R formula interface is required [@law2020guide].

```{r tableOverview}
xx <- list(
  MSstats = c("pd" = "Y", "rm" = "Y", "mem" = "y"),
  ROPECA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y", eb = "Y"),
  limma  = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y"),
  MSqRob2_rlm = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "?"),
  proDA = c("pd" = "Y", "rm" = "Y*", fd = "Y", int = "Y",eb = "Y", md = "Y"),
  prolfqua = c("pd" = "Y", "rm" = "Y", fd = "Y", int = "Y", mem = "Y",eb = "Y", md = "Y"))

bb <- data.frame(dplyr::bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, * - repeated measurements are modeled using a fixed effects. ? - the hurdle model was published but is not available in the msqrob2 package.")
```

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables to call various machine learning (ML) methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, what these packages have in common is that they fit linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented (OO) design, with R linear models as a core, where features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modelling of missing observations can be added.

@@

We also implemented functionality to streamline the benchmarking of the implemented methods. We use the Ionstar [@shen2018ionstar] dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, a high power of the tests is a relevant criterion to asses the modelling method. Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. However, the model must also provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments.

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base R[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

In addition we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes encapsulate the statistical modelling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (ContrastInterface), and there are various implementations allowing to perform differential expression analysis given linear or mixed effect models (Contrast), variance shrinkage (ContrastModerated), or impute contrasts in case of observations missing for an entire group of samples (ContrastSimpleImpute) can also be used to encapsulate and integrate differential expression analysis results of external tools such as proDA or to analyse data from protein interaction studies using SAINTexpress [@teo2014saintexpress].

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("lm", "linear model for peptides or proteins", "strategy_lm, Contrasts"),
    lmer = c("lmer" , "mixed effect models for peptides or proteins", "strategy_lmer, Contrasts"),
    gmi = c("group mean imputation", "impute missing group means", "ContrastsSimpleImpute"),
    ROPECA = c("ROPECA"," ", "ContrastsROPECA"),
    limma = c("limma" , "empirical Bayes variance shrinkage", "ContrastsModerated"),
    pic = c("SAINTexpress", "protein interaction scoring", "runSaint, ContrastsSaintExpress"),
    prodA = c("proDA", "dropout models", "strategy_proDA*, Contrasts_proDA*"), check.names =  FALSE)
xM <- data.frame(t(xM))

colnames(xM) <- c("id", "model", "prolfqua functions")
knitr::kable(tibble::as_tibble(xM[,1:3]), caption = "Models available in prolfqua. * in development",
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down")

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. By comparing the various dilutions, we obtain various effect sizes , e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, and hence 20 raw files in total. We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata].

## Data preprocessing for model comparison

The peptide abundances (from the _MaxQuant_ _peptide.txt_ file), were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. We also did not use imputation. Protein abundances are inferred from the peptide intensities using Tukey's median polish. We fitted the fixed effect linear models using protein abundances, the mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model we fitted to peptide intensities.

## Benchmark metrics

@@

In order to compute the confusion matrices based on the $p$-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the $p$-value is a function of the $t$-statistics and the degrees of freedom.

Valikangas [@valikangas2016systematic] and colleagues discuss and benchmark various methods of peptide or protein intensity normalization such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work we will be using a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median and instead of the standard deviation $\tilde{S}$ the median absolute deviation (mad):

$$
z = \frac{x - \bar{x}}{S} \sim \frac{x - \tilde{x}}{\tilde{S}}
$$

Because we need to estimate the protein fold-changes on the original scale, we have to multiply the $z$-score by the average variance of all the $N$ samples in the experiment.

$$
z' = z \times 1/N\sum_{i=1}^N S_i
$$

To apply this transformation, we need to estimate two parameters per sample. It works for experiments with thousands of proteins and experiments where only a few hundred proteins per sample are measured.
For the Ionstar dataset, we used the intensities of _H. sapiens_ proteins, whose concentrations do not change, to determine $\bar{x}$ and $S$ and then applied it to all the intensities (including _E. coli_) in the sample.

### Estimating differences between groups

Given a linear model $y = \beta X$, we can compute the difference between two groups $\beta_{c}$ by a linear combination $c$ of linear model parameters $\beta$. $c$ is vector with as many elements as there are coefficients in the linear model. If $c$ has $0$ for one or more of its rows, then the corresponding coefficient in $\beta$ is not involved in determining the contrast [@ph525xseries].

The difference estimate $\beta_c$, is given by:

\begin{eqnarray} 
\hat{\beta_{c}} &=& \sum l\beta 
\end{eqnarray}

with $X$ being the design matrix, $\beta$ the model parameters and $c$ an array of weights, defining the contrast (difference among conditions). The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are available in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

The package prolfqua provides function to determine the vector of $parameter$ weights $c$, from a contrast specification:

- `<contrast name>` = `<factor><level_1> - <factor><level_2>`

Furthermore, contrasts for subgroups also can be specified using the syntax:

- `<contrast name>` = `<factor_1><subgroup>:<factor_2><level_1>` - `<factor_1><subgroup>:<factor_2><level_2>`

where `<factor_x>` is the name of the explanatory variable, and `<level_x>` are group labels.

The following code shows an example where we specify two contrast: 
a) where we compare Cells of type CMP/MEP with cells of type HSC, and b) where we contrast therapy NO with therapy NU for the celltype CMP/MEP [@meier2021reduced].

```{r echo=TRUE}
Contrasts <- 
    c("CMP/MEPvsHSC" = "`CelltypeCMP/MEP` - `CelltypeHSC`",
   "NOvsHU" = "`class_therapyc.NO:CelltypeCMP/MEP` - `class_therapyp.HU:CelltypeCMP/MEP`")

m <- prolfqua::prolfqua_data('data_basicModel_p1807')
linfct <- prolfqua::linfct_from_model(m,as_list = FALSE)

prolfqua::linfct_matrix_contrasts(linfct, Contrasts )
```

Missing observations lead to different group sizes, that is, unbalanced designs. Linear and mixed effect models implemented in R can handle unbalanced designs. Therefore, as long as one observation per group is available, they will produce unbiased estimates, and no imputation is needed.
If there is no observation in a group, we assume that the protein is unobserved because the protein abundance is below the limit of detection (LOD) of the MS instrument. In this case, we will impute the mean using the protein abundance at the detection limit $A_{LOD}$.
We estimate the abundance at the detection limit using the abundances of the proteins observed only in one sample of a group of samples. Then, we compute the median of the distribution and use it as the group mean if a protein is absent in a treatment group.

When computing differences among two groups $a$ and $b$, we will use the group mean $\bar{a}$ or $\bar{b}$ estimated from the data, or if no observations are present we use $A_{LOD}$. If there are no observations in group $b$ then

$$
\Delta_{a,b} = 
\begin{cases}
\bar{a} - A_{LOD} & \text{if} ~~ \bar{a} > A_{LOD}\\
0  &\text{if} ~~ \bar{a} < A_{LOD}.
\end{cases}
$$

To estimate the variance, we assume that the variance of the protein is constant in all the groups and use the pooled variance based on all data:

Where, $n_g$ is the number of groups and $n$ is the number of observations.
If variance can not be estimated for a protein, we will use the median pooled variance of all the proteins.
